Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-10-06 pkg/glmertree/paper/treatment_interactions.Rnw
some small improvements in pages 1-6
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@996 255f4249-2263-4417-b1ea-8498779e27a7

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure. Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, as well accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of trees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of datasets, and trees without random effects in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4 and 33\%, respectively). Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal predictive accuracy (.60 on avergae), but better accuracy in detecting interactions. We illustrate the application of GLMM tree on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov15, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM tree.

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14), GLMM tree allows for taking into account the clustered structure of datasets. In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.

@@

We created generated a dataset representing observations on 150 participants in a randomized clinical trial. Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).

@@

@@

For example, let us take a global GLM to estimate the overall treatment effect in the motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper. Also, we assume a continuous response variable, employing a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented. As can be seen, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.

>>>>

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure. Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM trees and compare it with that of trees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of the simulated datasets, whereas trees without random effects only did so in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4\% and 33\%, respectively). Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal predictive accuracy (.60 on average), but better accuracy in detecting interactions. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are particularly useful for exploratory purposes, as they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. As such, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions. Several tree-based algorithms for the detection of treatment-subgroup interactions have already been developed (\citeNP{DussyDoov15, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08, SeibyZeil15}; see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, allowing users to select from a range of statistical methods, including trees.

In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM trees.

The GLMM tree algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}), which offers a flexible framework for subgroup detection. For example, GLM-based MOB has been applied to detect treatment-subgroup interactions for the treatment of depression \cite{DrieySmit16} and amyotrophic lateral sclerosis \cite{SeibyZeil15}. In contrast to other purely tree-based methods \cite<e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}), GLMM tree allows for taking into account the clustered structure of datasets. In contrast to previously suggested regression trees with random effects, GLMM tree allows for treatment effect estimation, with continuous as well as non-continuous response variables \cite<e.g., >{HajjyBell11,SelaySimo12}.

@@

We created a dataset representing observations on 150 participants in a randomized clinical trial. Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).

@@

See Figure~\ref{fig:example_glimmertree} for the GLMM tree fitted to the simulated data set which recovers the true underlying structure.

@@

For example, let us take a global GLM to estimate the overall treatment effect in the motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper. Also, for the continuous response variable in the motivating data set, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented. As already mentioned above, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.
